Towards Practical First-Order Model Counting

### 2024-04-19 doc/paper.tex
Section 3.1 (except for the proof)

\section{Introduction}

@@

\section{Preliminaries}

@@

\subsection{Logic}\label{sec:logic}

@@

\subsection{Algebra}\label{sec:algebra}

\todo[inline,caption={}]{
  \begin{itemize}
    \item say something about what $\expr{}$ can contain (function calls,
          various arithmetic operations, summations, non-negative integer
          constants, variables introduces by the function signature or
          summations)
    \item The definition of a function call contradicts the base case finding
          algorithm. Maybe rephrase it to only look at function calls from $f$
          to itself and assume:
          \begin{itemize}
            \item no mutual recursion,
            \item these calls in particular have this form
          \end{itemize}
  \end{itemize}
}

@@

\begin{definition}
  Given a function $f$ with signature $f(\mathbf{x})$, a \emph{function call} of
  $f$ is $f(\mathbf{y})$, where each $y_{i}$ is either $x_{i} - c_{i}$ or
  $c_{i}$ for some constant $c_{i} \in \mathbb{N}_{0}$.
\end{definition}

@@

\Cref{alg:main} outlines our overall approach for compiling a formula into a set
of equation that include the required base cases. In short, we first use the
knowledge compilation algorithm of the original
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23} to compile the formula into:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
After some algebraic simplification, $\mathcal{E}$ is passed to the
\FindBaseCases procedure that returns a set of base cases that we need to find
solutions for (described in \cref{sec:identifying}). For each base case
$f(\mathbf{x})$, we identify the formula associated with $f$ and simplify it
using the \Propagate procedure (described in \cref{sec:simplifying}). The
algorithm then recurses on these simplified formulas and adds the resulting base
case equations to $\mathcal{E}$. \Cref{example:overall} explains \cref{alg:main}
in more detail.

@@

  After \cref{line:first,line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain introduced by \Crane. Then
  \FindBaseCases identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both
  cases, \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$
  simplified by assuming that one of the domains is empty. In the first case, we
  recurse on the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$,
  where $S$ is a predicate introduced by Skolemization with weights
  $w^{+}(S) = 1$ and $w^{-}(S) = -1$. Hence, we get the base case
  $g(0, m) = 0^{m}$. In the case of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$}
  returns an empty formula, giving us $g(l, 0) = 1$.

  Note that these base cases overlap when $l = m = 0$ but are consistent with
  each other since $0^{0} = 1$. More generally, let $\phi$ be a formula with two
  domains $\Gamma$ and $\Delta$, and let $n, m \in \mathbb{N}_{0}$. Then the
  model count of \Propagate{$\phi$, $\Delta$, $n$} assuming $|\Gamma| = m$ is
  the same as the model count of \Propagate{$\phi$, $\Gamma$, $m$} assuming
  $|\Delta| = n$.
\end{example}

\todo[inline]{This last claim sounds like something that should be proven.}

@@

  $\mathcal{B} \gets \emptyset$\;
  \ForEach{equation $(\ldots = \expr{}) \in \mathcal{E}$}{
    \ForEach{function call $f(\mathbf{y}) \in \expr{}$}{\label{line:functioncall}
      let $f(\mathbf{x})$ be the signature of $f$\;
      \ForEach{$y_{i} \in \mathbf{y}$} {
        \uIf{$y_{i} \in \mathbb{N}_{0}$}{
          $\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto y_{i}] \,\}$\;
        }
        \ElseIf{$y_{i} = x_{i} - c_{i}$}{
          \For{$j \gets 0$ \KwTo $c_{i} - 1$}{\label{line:lim}
            $\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto j] \,\}$\;\label{line:insert}
          }
        }
      }
    }
  }
\end{algorithm}

\todo[inline,caption={}]{
  \begin{itemize}
    \item introduce the Iverson bracket (it's only used in this subsection)
    \item \Cref{alg1} is now fixed and supersedes the description below (and the
          code!)
    \item NOTE: exactly one of the if and else-if branches always applies
  \end{itemize}
}

The algorithm is described as \cref{alg1}. We know that if, say, on the RHS of
all equations, the domain size appears as $m - c_1, m - c_2, \dots, m - c_k$,
then finding $f(0, x_1, x_2, \dots)$, $f(1, x_1, x_2, \dots)$,
$\dots f(m_0, x_1, x_2, \dots)$ for every function $f$, where
$m_0 = \max(c_1, c_2, \dots c_k) - 1$ forms a sufficient set of base cases.
Hence, in order to do the same efficiently, we can take that domain for which
$m_0$ is the minimum, i.e. $\argmin(\max(c_1, c_2, \dots c_k))$. Ideally, we
should calculate the base cases by finding the base cases up to
$\max(c_{1}, c_{2}, \dots) - 1$. However, currently only empty and singleton
domains are supported.

First, expand the summations in each equation. Here we expand the summations of
the form: $\sum_{x=0}^{x_{1}} \expr{} \cdot [a \le x < b]$ or similar
inequalities where $x$ is bounded by constants and $a$ and $b$ are constants, by
substituting the value of $x$ from $a$ to $b-1$. For example, we replace
$\sum_{x=0}^{x_1} \binom{x_{1}}{x} f(x_1 - x) \cdot [0 \le x < 2]$ by
$\binom{x_{1}}{0} f(x_1) + \binom{x_{1}}{1} f(x_1-1)$.

\begin{itemize}
  \item \cref{line:insert} for $f(\mathbf{x}) = f(y, z)$, $x_{i} = y$, and
        $n = 0$, add $f(y, z)[y \mapsto 0] = f(0, z)$ to $\mathcal{B}$.
\end{itemize}

\begin{theorem}
  Under the following assumptions, \cref{alg1} is guaranteed to return a
  sufficient set of base cases:
  \begin{itemize}
    \item there is no mutual recursion
    \item each function $f$ has exactly one equation with $f$ on the LHS;
    \item in the recursive definition of function $f(x_{1},\dots,x_{n})$, the
          $i$-th argument of each call to $f$ on the RHS is of the form
          $x_{i} - c_{i}$, where:
          \begin{itemize}
            \item $c_{i} \ge 0$ for all $i$, and
            \item $c_{i} > 0$ for at least one $i$.
          \end{itemize}
  \end{itemize}
\end{theorem}

\begin{example}
  For an equation $f(m, n) = 2 \times f(m-1, n)$, \cref{alg1} returns $f(0, n)$.
\end{example}

@@

\section{Smoothing}

@@

\subsection{Smoothing for Base Cases}\label{sec:smoothingbase}

@@

\subsection{Smoothing the FCG}

@@

\subsection{Stage 1: Propagating Unit Clauses `Upwards'}

@@

\subsection{Stage 2: Adding Smoothing Nodes}

@@

\section{Generating C++ Code}\label{sec:cpp}

@@

\section{Experimental Evaluation}

@@

\section{Conclusion}

>>>>

\section{Introduction (TODO)}

@@

\section{Preliminaries (TODO)}

@@

\subsection{Logic (TODO)}\label{sec:logic}

@@

\subsection{Algebra (TODO)}\label{sec:algebra}

@@

@@

\Cref{alg:main} outlines our overall approach for compiling a formula into a set
of equation that include the required base cases. In short, we first use the
knowledge compilation algorithm of the original
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23} to compile the formula into:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
After some algebraic simplification, $\mathcal{E}$ is passed to the
\FindBaseCases procedure (described in \cref{sec:identifying}) that returns a
set of base cases that we need to find solutions for. For each base case
$f(\mathbf{x})$, we identify the formula associated with $f$ and simplify it
using the \Propagate procedure (described in \cref{sec:simplifying}). The
algorithm then recurses on these simplified formulas and adds the resulting base
case equations to $\mathcal{E}$. \Cref{example:overall} explains \cref{alg:main}
in more detail.

@@

  After \cref{line:first,line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain introduced by \Crane. Then
  \FindBaseCases identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both
  cases, \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$
  simplified by assuming that one of the domains is empty. In the first case, we
  recurse on the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$,
  where $S$ is a predicate introduced by Skolemization with weights
  $w^{+}(S) = 1$ and $w^{-}(S) = -1$. Hence, we get the base case
  $g(0, m) = 0^{m}$. In the case of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$}
  returns an empty formula, giving us $g(l, 0) = 1$.
\end{example}

Note that these base cases overlap when $l = m = 0$ but are consistent with each
other since $0^{0} = 1$. More generally, let $\phi$ be a formula with two
domains $\Gamma$ and $\Delta$, and let $n, m \in \mathbb{N}_{0}$. Then the model
count of \Propagate{$\phi$, $\Delta$, $n$} assuming $|\Gamma| = m$ is the same
as the model count of \Propagate{$\phi$, $\Gamma$, $m$} assuming $|\Delta| = n$.

\todo[inline]{Can I turn the above into a theorem?}

Finally, we note that the \Simplify procedure plays a crucial role in
simplifying a common algebraic pattern $\sum_{m=0}^{n}[a \le m \le b] f(m)$.
Here:
\begin{enumerate*}[label=(\roman*)]
  \item $n$ is a variable,
  \item $a, b \in \mathbb{N}_{0}$ are constants,
  \item $f$ is an expression that may depend on $m$, and
  \item $[a \le m \le b] =
  \begin{cases}
    1 & \text{if $a \le m \le b$} \\
    0 & \text{otherwise}
  \end{cases}$ is the Iverson bracket.
\end{enumerate*}
\Simplify transforms this pattern into
$f(a) + f(a+1) + \cdots + f(\min\{\, n, b \,\})$. For instance, in the case of
\cref{example:overall}, \Simplify transforms
$g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)$ into the
simpler form above.

\todo[inline]{A concluding sentence that introduces the remaining subsections?}

@@

  $\mathcal{B} \gets \emptyset$\;
  \ForEach{equation $(f(\mathbf{x}) = \expr{}) \in \mathcal{E}$}{
    \ForEach{function call $f(\mathbf{y}) \in \expr{}$}{\label{line:functioncall}
      \ForEach{$y_{i} \in \mathbf{y}$} {
        \uIf{$y_{i} \in \mathbb{N}_{0}$}{
          $\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto y_{i}] \,\}$\;
        }
        \ElseIf{$y_{i} = x_{i} - c_{i}$ for some $c_{i} \in \mathbb{N}_{0}$}{
          \For{$j \gets 0$ \KwTo $c_{i} - 1$}{\label{line:lim}
            $\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto j] \,\}$\;\label{line:insert}
          }
        }
      }
    }
  }
\end{algorithm}

\Cref{alg1} summarises the implementation of \FindBaseCases. For each recursive
call from a function $f$ to itself, we consider two types of arguments:
\begin{enumerate*}[label=(\roman*)]
  \item constants and
  \item arguments of the form $x_{i} - c_{i}$, where $c_{i} \in \mathbb{N}_{0}$
  is a constant, and $x_{i}$ is the $i$-th argument of the signature of $f$.
\end{enumerate*}
In the former case, we consider a base case specifically for that constant. In
the latter case, we consider a base case for constants from zero up to (but not
including) $c_{i}$. \Cref{thm:halting} below motivates this approach.

\begin{example}
  Consider the recursive function $g$ from \cref{example:overall}.
  \FindBaseCases{$\mathcal{E}$} iterates over two function calls: $g(l-1, m)$
  and $g(l-1, m-1)$. The former produces the base case $g(0, m)$, while the
  latter produces both $g(0, m)$ and $g(l, 0)$.
\end{example}

\todo[inline]{Prove (lots of notes, handwritten and elsewhere)}

\begin{theorem}[Halting]\label{thm:halting}
  Let us assume the following about the input set of equations $\mathcal{E}$:
  \begin{enumerate}
    \item Each function $f$ has exactly one equation with $f$ on the LHS.\@ We
          call this equation the \emph{definition} of $f$.
    \item There exists a topological ordering of all functions ${(f_{i})}_{i}$
          such that the definition of $f_{i}$ does not contain function calls to
          $f_{j}$ with $j > i$.\footnote{This condition excludes the possibility
          of mutual recursion and similar cyclic scenarios and is akin to
          stratified logic programs~\cite{DBLP:books/sp/Lloyd87}.}
    \item For every equation $(f(\mathbf{x}) = \expr) \in \mathcal{E}$, every
          recursive function call $f(\mathbf{y}) \in \expr$ satisfies the
          following:
          \begin{itemize}
            \item each $y_{i}$ is either $x_{i} - c_{i}$ or $c_{i}$ for some
                  constant $c_{i} \in \mathbb{N}_{0}$;
            \item there exists $i$ such that $y_{i} = x_{i} - c_{i}$ for some
                  $c_{i} > 0$.
          \end{itemize}
  \end{enumerate}
  Then the set of base cases $\mathcal{B}$ returned by
  \FindBaseCases{$\mathcal{E}$} is \emph{sufficient} for $\mathcal{E}$ in the
  following sense. The evaluation of any function call with non-negative
  arguments halts as long as the evaluation of base cases is prioritised over
  the corresponding recursive definitions.\footnote{Recall that, as previously
    discussed, the order in which base cases are considered is immaterial.}
\end{theorem}

@@

\section{Smoothing (TODO)}

@@

\subsection{Smoothing for Base Cases (TODO)}\label{sec:smoothingbase}

@@

\subsection{Smoothing the FCG (TODO)}

@@

\subsection{Stage 1: Propagating Unit Clauses `Upwards' (TODO)}

@@

\subsection{Stage 2: Adding Smoothing Nodes (TODO)}

@@

\section{Generating C++ Code (TODO)}\label{sec:cpp}

@@

\section{Experimental Evaluation (TODO)}

@@

\section{Conclusion (TODO)}
